Software architectures to integrate workflow engines in science gateways

### 2016-10-11 answers.tex
More answers to the reviewers.

\documentclass[a4]{letter}
\usepackage{graphicx}
\usepackage{color}
\usepackage{hyperref}
\name{Tristan Glatard}
\signature{\includegraphics[width=4cm]{signature.png}}
\date{October 16th, 2016}

\newenvironment{review}%          environment name
{\textbf{Reviewer comment:}\begin{quote}}% begin code
{\end{quote}}%  
\newcommand{\answer}[1]{\textbf{Answer:} #1}

@@

\begin{letter}{}

\opening{Dear Editors,}

We would like to thank you for offering us the opportunity to revise
and re-submit our paper entitled ``Software architectures to integrate
workflow engines in science gateways'' to Future Generation Computer
Systems.

The remainder of this letter contains our answer to the 3 reviews that
we received following the evaluation process. A revised version of the
manuscript follows, where corrected parts appear in blue for a better
identification.

We thank you again for your time and consideration, and we remain available for any further information you may need.

\vspace{0.5cm}

Sincerely yours,

\vspace{1cm}

\closing{Tristan Glatard, PhD.\\On behalf of the authors.}

\newpage

\textbf{Reviewer \#1}

@@

  The fact that the patterns may be mixed in real implementations is,
  in our opinion, another motivation for our abstraction effort as it
  helps ``untangling'' the various architectural concepts that may be
  present in a real system. This is particularly true for the abstract
  architectures (nested workflows and workflow conversion) that may be
  instantiated in various ways. Our abstractions highlight that other
  instantiation models may be worth designing, for instance nested
  workflows with task encapsulation, as mentioned in Section 4.2. in
  the Discussion.

@@

We definitely agree with the reviewer that interaction (g) is
  a tricky one that may lead to critical errors. We added the
  following sentence to the paragraph on interaction (g) in Section
  4.2. of the discussion:

@@

\answer{As discussed previously, we share the reviewer's concerns
  about the feasibility of workflow conversion (see comments on
  interaction (g) in Section 4.2). The performance of workflow
  conversion is strongly related to the availability of a validated
  interaction (g), which is very dependent on the particular workflow
  engines under consideration. We believe that the issue raised by the
  reviewer basically comes down to the weight given to software
  components and interactions in the analysis. However, using
  different weights for different components or interactions would
  open an infinite number of configurations for discussion. For
  instance, in some cases, the workflow engine already comes with a
  workflow service which therefore doesn't have to be included. To
  address this issue, we shared a spreadsheet at
  \url{https://frama.link/LrPRsCQp} that allows editing the following
  weights:
  \begin{itemize}
  \item Weight given to software components and interactions in
  ``integration'', ``extensibility'' and ``robustness'' criteria.
  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
  ``extensibility'').
  \item Weight given to criteria in overall evaluation of Table 3.
\end{itemize}
The paper references this spreadsheet wherever equal weights are
assumed. The readers and reviewers are invited to rely on this
spreadsheet for any issue related to the importance given to a
specific component, interaction, metric or criterion in the
analysis. In the paper, we assumed that all weights equal 1, which we
believe is a reasonable but limited choice.
}

@@

\answer{The reviewer is right that re-purposing software components
  and interactions can lead to revisit the evaluation. We updated the following sentence in section 4.2. to better acknowledge this point:}

@@

\answer{Yes, as for any other metric, high values indicate worse
performance. This is explained at the beginning of section 3:
``Criteria break down into specific met- rics where lower value
indicates better performance.''}

@@

@@

\textbf{Reviewer \#2}

@@

@@

@@

@@

@@

@@

@@

@@

\textbf{Reviewer \#3}

@@

@@

@@

\end{letter}
\end{document}

>>>>

\documentclass[a4]{article}
\usepackage{graphicx}
\usepackage{color}
\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links

\usepackage{hyperref}
\newenvironment{review}%          environment name
{\textbf{Reviewer comment:}\begin{quote}}% begin code
{\end{quote}}%  
\newcommand{\answer}[1]{\textbf{Answer:} #1}
\newcommand{\todo}[1]{\color{red}#1\color{black}}

@@

\section{Reviewer \#1}

@@

  The fact that the patterns may be mixed in real implementations is,
  in our opinion, another motivation for our abstraction effort as it
  helps ``untangling'' the various architectural concepts that may be
  present in a real system. This is particularly true for the abstract
  architectures (nested workflows and workflow conversion) that may be
  instantiated in various ways. Our abstractions highlight that other
  instantiation models may be worth designing, for instance nested
  workflows with task encapsulation, as mentioned in Section 5.2. in
  the Discussion.

@@

We definitely agree with the reviewer that interaction (g) is
  a tricky one that may lead to critical errors. We added the
  following sentence to the paragraph on interaction (g) in Section
  5.2. of the discussion:

@@

\answer{As discussed previously, we share the reviewer's concerns
  about the feasibility of workflow conversion (see comments on
  interaction (g) in Section 5.2). The performance of workflow
  conversion is strongly related to the availability of a validated
  interaction (g), which is very dependent on the particular workflow
  engines under consideration. We believe that the issue raised by the
  reviewer basically comes down to the weight given to software
  components and interactions in the analysis. However, using
  different weights for different components or interactions would
  open an infinite number of configurations for discussion. For
  instance, in some cases, the workflow engine already comes with a
  workflow service which therefore doesn't have to be included. To
  address this issue, we shared a spreadsheet at
  \url{https://frama.link/LrPRsCQp} that allows editing the following
  weights:
  \begin{itemize}
  \item Weight given to software components and interactions in
  ``integration'', ``extensibility'' and ``robustness'' criteria.
  \item Weight given to metrics in a criterion (e.g., E$_1$ vs. E$_2$ in
  ``extensibility'').
  \item Weight given to criteria in overall evaluation of Table 3.
\end{itemize}
The paper references this spreadsheet wherever equal weights are
assumed. The readers and reviewers are invited to rely on this
spreadsheet for any issue related to the importance given to a
specific component, interaction, metric or criterion in the
analysis. In the paper, we assumed that all weights equal 1, which we
believe is a reasonable but limited choice.
}

@@

\answer{The reviewer is right that re-purposing software components
  and interactions can lead to revisit the evaluation. We updated the following sentence in section 5.2. to better acknowledge this point:}

@@

\answer{Yes, as for any other metric, high values indicate worse
performance. This is explained at the beginning of section 3:
``Criteria break down into specific metrics where lower value
indicates better performance.''}

@@

\todo{think of this}

@@

\answer{This was done as suggested.}

\section{Reviewer \#2}

@@

\todo{just remove CSGF, Airvata and pgrade from the table, keeping the
  exponents. But we don't have examples of gateways using Airvata,
  still need to be found.}

@@

\answer{The task submission framework is part of the infrastructure
  and its nature doesn't influence the architecture used to integrate
  workflow engines in the science gateway. Formally, it is part of
  interaction \texttt{b} (see definition in 2.1).

  If the ``entire workflow'', i.e. the complete control and data flow,
  is submitted to the infrastructure through interaction \texttt{b} as
  a single task that will manage the execution and potentially submit
  sub-tasks, then this is a task encapsulation architecture. In
  particular, a dedicated engine instance can easily be started for
  every workflow execution.

  If only the workflow tasks are submitted to the infrastructure
  through interaction \texttt{b}, and the main workflow logic is
  controlled from within the science gateway, then this is clearly a
  tight integration.

  The reviewer's use case, however, suggest that the workflow engine
  is split in two parts: (1) a high-level part integrated in the
  science gateway, which submits ``entire workflows'' to the
  infrastructure; (2) a lower-level part, executed on the
  infrastructure, which executes such ``entire workflows''. It seems
  to us that this matches the definition of a nested workflow model,
  where the parent engine is integrated in the science gateway, and
  the child engine runs on the infrastructure. This is another
  instantiation of the nested workflow abstract model, which is not
  evaluated in the paper as mentioned in section 5.2: "\emph{The
    abstract nested workflows and workflow conversion patterns were
    instantiated with service invocation so that they can be analyzed
    in the same framework as the other architectures. Other types of
    instantiation, for instance nested workflows with task
    encapsulation, could also be envisaged.}". It sounds like the
  architecture described by the reviewer is a nested workflow pattern
  instantiated with a tight integration (parent engine) and task
  encapsulation (child engine).

  We found this scenario interesting and we evaluated it in our
  framework. The analysis is reported in the new section 6 of the
  paper.}

@@

\todo{Don't divide by the total number of interactions because then a
  given interaction could have a different value depending on the
  architecture. Refer to spreadsheet where weights can be adjusted.}

@@

\todo{OK but these components are not specific to workflow
  execution. So it's out of the definition. Motivation for this definition: these components are here anyway. If the SG fails, then the
  workflows can't even be started. If the infrastructure is down, same. }

@@

\todo{no, because c1 is not red.}

@@

\todo{I agree it's boring but it's also important to have it clearly written.}

@@

\todo{what???}

@@

\section{Reviewer \#3}

@@

\todo{answer metric by metric, focusing on E1 and E4. For E4 it's clear. For E1, }

@@

\answer{complexity is used as a measure of reliability}

@@

\end{document}

### 2016-10-11 paper.tex
More answers to the reviewers.

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{pdfcomment}

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}
\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}

@@

\correction{The reminder of this section provides}
 background information and definitions of workflow engines,
science gateways and infrastructures. In
Section~\ref{sec:architectures}, we describe six architectures that
cover the main patterns that we encountered in the various systems
that we studied during the last few years. The architectures are
described within a consistent framework that underlines the functional
interactions between their main software components. In
Section~\ref{sec:evaluation}, the different architectures are
evaluated with metrics measuring integration complexity, robustness,
extensibility, scalability and other specific features. These metrics
are specifically designed to measure the ability of the architectures
to address workflow-related issues commonly encountered in science
gateways. We finally compare the architectures and highlight
recommendations for science gateway architects and developers.

@@

\section{Evaluation}

@@

The criteria and metrics are
explained hereafter.

@@

\subsection{Evaluation metrics}

@@

The architectures described in Figure~\ref{fig:architectures} are
evaluated along these metrics in the remainder of this Section.

@@

\paragraph{Integration complexity} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires developing interaction \texttt{e} and  installing the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
in the architecture shares the same mechanism as version updates of
other tasks, which requires an update on the
infrastructure (\texttt{E$_2$}=1).  New workflows are integrated by
creating a new task in the science gateway through interactions
\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
infrastructure requires updating interactions \texttt{a$_1$},
\texttt{b} and \texttt{c} in the science gateway (\texttt{E$_4$}=3).

@@

@@

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. \correction{Weights can be adjusted in the
  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
  sheet ``Metrics'') to produce a custom version of
  Table~\ref{table:overall} if required.}

@@

The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \texttt{c}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.

>>>>

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pbox}
\usepackage{pdfcomment}
\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
%\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
%\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
%\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
%\newcommand{\closedanswerednote}[6]{}
\newcommand{\correction}[1]{\color{blue}#1\color{black}\xspace}

@@

\correction{The reminder of this section provides} background
information and definitions of workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe six
architectures that cover the main patterns that we encountered in the
various systems that we studied during the last few years. The
architectures are described within a consistent framework that
underlines the functional interactions between their main software
components. \correction{In Section~\ref{sec:metrics}, we introduce
  metrics that measure integration complexity, robustness,
  extensibility, scalability and other specific features. These
  metrics are specifically designed to measure the ability of the
  architectures to address workflow-related issues commonly
  encountered in science gateways. In Section~\ref{sec:evaluation}, we
  evaluate the different architectures based on the proposed metrics.}
We finally compare the architectures and highlight recommendations for
science gateway architects and developers.

@@

\section{\correction{Metrics}}

\label{sec:metrics}

@@

@@

@@

\section{\correction{Evaluation}}

The architectures described in Figure~\ref{fig:architectures} are
evaluated \correction{hereafter with the proposed metrics}.

@@

\paragraph{Integration complexity} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \correction{\texttt{c$_1$}} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Extensibility} Integrating a new type of workflow engine
requires developing interaction \texttt{e} and  installing the engine
on the infrastructure (\texttt{E$_1$}=2). Updating an engine version
in the architecture shares the same mechanism as version updates of
other tasks, which requires an update on the
infrastructure (\texttt{E$_2$}=1).  New workflows are integrated by
creating a new task in the science gateway through interactions
\texttt{a$_1$} and \texttt{a$_2$} (\texttt{E$_3$}=2). Adding a new
infrastructure requires updating interactions \texttt{a$_1$},
\texttt{b} and \correction{\texttt{c$_1$}} in the science gateway (\texttt{E$_4$}=3).

@@

\correction{\section{Real systems}}

\label{sec:real systems}

\correction{\section{Application to systems design}}
\label{sec:application}

\correction{To illustrate how our evaluation framework can be applied to the
design of new systems, let's consider a hypothetical system where the
workflow engine is split in two parts: (1) a high-level part
integrated in the science gateway, which submits sub-workflows to the
infrastructure; (2) a lower-level part, executed on the
infrastructure, which executes the sub-workflows. This scenario was
suggested by Reviewer \#2 of the paper after the evaluation framework
had been designed.

We model such a system using a nested workflow architecture where the
parent engine is tightly integrated in the science gateway, and the
child engine is integrated through task encapsulation.
Figure~\ref{fig:nested-3} shows the graphical representation of the
system. From this representation we find that the sub-tasks submitted
by the child engine through interaction \texttt{e} should be handled
by the parent engine rather than by the science gateway. Indeed,
handling the tasks directly in the science gateway would lead to
create a new specific interaction, named, e.g., \texttt{b$_1$}, while
\texttt{b} could be reused. We also find that interaction \texttt{e}
allows running the sub-tasks on the infrastructure supported by the
parent engine with no further development effort required in the child
engine.

The system evaluation is presented on
Table~\ref{table:evaluation-2}. The detailed calculations are
available in sheet ``Metrics with reviewer 2's architecture'' of the public spreadsheet.

We find that the studied system supports nested workflows with a much
better performance than the nested workflow model instantiated with
service invocation: the integration score is improved by a factor of
2, the robustness is improved by a factor of 2.6, the extensibility is
substantially improved, the scalability is not penalized, and the
specific features are slightly improved.}

\begin{figure}
  \centering
  \def\svgwidth{0.5\columnwidth}
  \input{figures/nested-3.pdf_tex}
  \caption{\correction{Nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
  \label{fig:nested-3}
\end{figure}

\begin{table}
\footnotesize
\centering
\begin{tabular}{rc}
                                     & \textbf{Nested 1}\\
\cellcolor[HTML]{EEEEEE}\textbf{Integration complexity}&\cellcolor[HTML]{EEEEEE}\\
  Total components -- \texttt{I$_1$} & 2\\
Total interactions -- \texttt{I$_2$} & 6\\
\textbf{Total} (total software pieces) & \textbf{8}\\
\cellcolor[HTML]{EEEEEE}\textbf{Robustness}&\cellcolor[HTML]{EEEEEE}\\
Specific components -- \texttt{R$_1$} & 0\\
  Specific interactions -- \texttt{R$_2$} & 3\\
  \textbf{Total} (execution software pieces)& \textbf{3}\\
\cellcolor[HTML]{EEEEEE}\textbf{Extensibility}&\cellcolor[HTML]{EEEEEE}\\
  New engine type -- \texttt{E$_1$}  & 3\\
New engine version -- \texttt{E$_2$} & 1\\
  New workflow -- \texttt{E$_3$}     & 2\\
New infrastructure -- \texttt{E$_4$} & 4\\
  \textbf{Total} (difficulty to extend) & \textbf{10}\\
\cellcolor[HTML]{EEEEEE}\textbf{Scalability}& \cellcolor[HTML]{EEEEEE}\\
Multiple engine instances -- \texttt{S$_1$}& 1\\
Distributed engines -- \texttt{S$_2$}& 0\\
Task scheduling -- \texttt{S$_3$}    & 1\\
\textbf{Total} (scalability issues)  & \textbf{2}\\
\cellcolor[HTML]{EEEEEE}\textbf{Specific features}&\cellcolor[HTML]{EEEEEE}\\
  Meta-workflow  -- \texttt{O$_1$}    & 0\\
  Fine-grained debugging -- \texttt{O$_2$}  & 0.5\\
  \textbf{Total} (missing features) & \textbf{0.5}\\
\end{tabular}
\caption{\correction{Evaluation of the nested workflow architecture instantiated with tight integration (parent engine) and task encapsulation (child engine).}}
\label{table:evaluation-2}
\end{table}

@@

Table~\ref{table:overall} \todo{update with new architecture?} provides an overall comparison between
architectures, based on the metrics in Table~\ref{table:evaluation}.
Note that this analysis is only meant for illustration purposes, since
it assumes that all criteria have equal weights while real systems
would favor some of them. \correction{Weights can be adjusted in the
  public spreadsheet at \url{https://frama.link/LrPRsCQp} (column B of
  sheet ``Metrics'') to produce a custom version of
  Table~\ref{table:overall} if required.}

@@

The particular technical or historical context of a science gateway
project may also influence the choice of an architecture to
integrate workflow engines. For instance, many workflow engines are
already available as web services, which tends to favor service
invocation (\correction{in particular when interactions can be re-purposed}), and other science gateways may have strongly tested task
and data control features (interactions \texttt{b} and \correction{\texttt{c$_1$}}),
which would favor task encapsulation. Similarly, adding a new type of engine
may facilitate integration of a new infrastructure when interactions
\texttt{b} and \texttt{c$_2$} are already available. The migration
cost between architectures has been ignored as well.
